---
title: "Tutorial 7.2 - Advanced Visualization: Interactive Graphs and Plotly"

author: "Zachary Binney, PhD"
date: "January 2021"

output: 
  learnr::tutorial:
    css: css/zb_style.css
    progressive: TRUE
runtime: shiny_prerendered
description: "Maps"
---

```{r hidden, include = FALSE}
pacman::p_load(learnr)

```


## Intro

You've learned a lot about how to make static plots you can show to people. These are nice because they allow you to perfectly control what people see. But their major limitation is the same: they require you to choose ahead of time exactly what people see. 

Sometimes you may want your readers to be able to interact with your data visualizations (or tables). Maybe you want, for example, to let someone scroll over a specific point or bar and get more information about it. Or [here](https://stateofchildhoodobesity.org/adult-obesity/), where hovering over a state highlights a line showing its obesity trend over time. This looks very slick and modern! Much more impressive than a simple static graph.

These sort of interactive data presentations are already very common in, for example, in business analytics dashboards like those made in Tableau. They're becoming more common every day in other areas, too.

### Packages

As usual we need to load a few packages. We need `tidyverse` because we're still going to be using many of its functions for data manipulation and plotting. And since this Tutorial is all about interactive plots with `plotly`, hopefully it's no surprise we need to load that, as well.

```{r setup, message = FALSE}
# Load (and, if necessary, first install) packages
pacman::p_load(tidyverse, plotly, gapminder, RColorBrewer)

```

### Motivating Example

Do you remember Homework 2 from QTM 150? One of the things I asked you to do with the `midwest` data frame was to identify the name of an outlier county in your scatterplot of county-level college attendance and poverty. To find the county name with the skills you had then you actually had to open the data frame and go manually searching for the counties matching the values of the outlier points. It was a massive pain! Later you learned skills like `filter()` that may have helped you do this in a script in a more automated way, but allow me to show you an even better solution with `plotly`.

Here's a static version of our `midwest` college-poverty plot:

```{r plotly_motiv1}
ggplot(data = midwest, mapping = aes(x = percollege, y = percbelowpoverty)) +
    geom_point() +
    geom_smooth() +
    labs(x = "Percent College Educated", y = "Percent Below Poverty Line")
```
And here's an interactive version we could build with `plotly`:

```{r plotly_motiv2}
plot_ly(midwest, x = ~percollege, y = ~percbelowpoverty) %>%
  add_markers(color = I("black"),
               text = ~paste0('County: ', county, ', ', state))
```
Now it's much easier to find the names of interesting outlier counties for further research, isn't it?

## `plotly` Overview

Unfortunately, as you probably noticed from the above example `plot_ly()` uses a somewhat different syntax from `ggplot()`. So we'll have to learn a new "grammar" for producing interactive plots. Fortunately there is some overlap, so learning `plot_ly()` should be substantially easier than your first time with `ggplot()`. Some of the major differences include:

* `plot_ly()` uses the pipe `%>%` all the way through, never the `+` operator to add more layers.

* `plot_ly()` uses the term **trace** where `ggplot()` would use **geom** or **layer**

    + You create new layers/traces on charts using `add_XX()` rather than `geom_XX()`. For example, to add a scatterplot player you use `add_markers()` rather than `geom_point()`

* If you don't tell it an explicit chart type, `plot_ly()` will try to *guess* the type of plot you want based on the type of data provided or other default settings. Note you should probably just always tell it a chart type to avoid this confusion!

* When specifying **aesthetics** you no longer need to place them inside `aes()`, but you *do* need to precede the variable name with `~`

* `plot_ly()` uses `layout()` rather than `theme()` to customize many elements of a plot such as the legend

* `plot_ly()` doesn't have an explicit `group =` aesthetic. You have to either use `group_by()` on a data frame *before* calling `plot_ly()` *OR* use a grouping aesthetic like `color =` so `plot_ly()` will implicitly create the groups for you (SEE the section on line plots below)

* `plot_ly()` has different defaults for its `geom_bar()` equivalent

But there are also some big similarities:

* `plot_ly()` also paints graphs in layers (it just calls them traces), which you add similarly

* Global and local aesthetics work quite similarly, and you add colors and text in much the same way as with `ggplot()`

## Our First Interactive Plot: Scatterplots

As noted above, like `plotly` also draws plots in layers. It starts with an empty canvas:

```{r plotly_empty, exercise = TRUE}
# Empty plotly graph
plot_ly()

```
Note that as mentioned above, since `plot_ly()` isn't given a plot type it tries to figure one out and settles on a scatterplot (defined as a "'scatter' **trace**" with the **mode** "markers").

We can see that it's actually trying to do this by feeding it a simple `x` and `y` value:

```{r plotly_one_point, exercise = TRUE}
# Slightly less empty plotly graph
plot_ly(x = 2, y = 2)

```
We could also ask for this more explicitly using the code below:

```{r plotly_one_point_explicit, exercise = TRUE}
# Slightly less empty plotly graph
plot_ly() %>% 
  add_markers(x = 2, y = 2)

```
Note two things: 1. We used a pipe `%>%` instead of `+` to add a layer; 2. We used `add_markers()` to tell `plot_ly()` to "add" a trace of type "marker" (a scatterplot, in `plot_ly()` speak).

Let's do a more useful plot: a scatterplot of `gdpPercap` and `lifeExp` from `gapminder`:

```{r plotly_scatter_gap, exercise = TRUE}
# Gapminder scatterplot
plot_ly(data = gapminder) %>% # First specify data = <DATA FRAME NAME>
  
  add_markers(x = ~ gdpPercap, # Specify the x aesthetic maps to ~ <VARIABLE NAME> (in this case, gdpPercap)
              
              y = ~ lifeExp)  # Specify the y aesthetic maps to ~ <VARIABLE NAME> (in this case, lifeExp)

```

#### Global vs. Local Aesthetics

Note the code above uses the "local aesthetics" method you used in `ggplot()` to specify the X and Y aesthetics only for that specific `add_markers()` **trace**. What if you wanted to map these *globally* for all layers of the graph (that is, on all layers have X be `gdpPercap` and Y be `lifeExp`), though?

<mark>Challenge:</mark> Alter the above code to produce the **same plot** but with X and Y defined globally for all layers rather than specific to a single layer. HINT: It works just like in `ggplot()`.

#### Mapping Additional Variables

What if we wanted to color the points by continent? Can you figure out how to do that? Once again, it's pretty much just like in `ggplot()`!

<mark>Challenge:</mark> Try to further alter the above code to color the plot's points by continent. Do this *before looking below*.

### Changing Colors in `plot_ly()`

One of the super nice things about `plot_ly()` is it has those `RColorBrewer` schemes built in! If you don't remember these schemes review that section in Tutorial 1.3 or check out the palettes below:

```{r cbpals, fig.height = 10}

display.brewer.all()

```

To change to one that's appropriate for your graph you simply specify the desired palette in the `colors =` argument:

```{r plotly_scatter_gap_colors, exercise = TRUE}
# Gapminder scatterplot
plot_ly(data = gapminder,
        x = ~ gdpPercap,
        y = ~ lifeExp,
        color = ~ continent,
        colors = "Dark2") %>% # Use the "Dark2" ColorBrewer palette
  add_markers()

```
Try playing around with palettes until you find one you like.

As indicated on the `plot_ly()` help page (<mark>Challenge:</mark> Do you remember how to pull that up in RStudio? Try it now!) the `colors =` argument also lets you specify a color ramp or a manual vector of colors (either names like `"gray"` or RBG vectors like `rgb(195, 195, 195)`. For more details on how to specify these I will refer you to [this post](http://rstudio-pubs-static.s3.amazonaws.com/448200_6bb02977b4c04e0da508ac0131f71d48.html).

## Interactive Line Plots and Specifying Traces

Let's try and make a line plot of how `gdpPercap` (adjusted for inflation) changes over time in every country in the `gapminder` data frame:

```{r plotly_line, exercise = TRUE}
# Line plot of per capita GDP by country
plot_ly(data = gapminder,
        x = ~ year,
        y = ~ gdpPercap,
        color = ~ country) %>% 
  add_lines() # This is what you use to make a line plot

```
With so many countries we get a very messy-looking splash of colors (the default `Set2` palette split into 142 different shades rather than the designed 8) and an absurdly long legend. Let's make three changes to the above: 1. Make all lines the same color (the default muted blue), 2. Remove the legend using `layout()`, and 3. Make the lines thinner so they're more easily differentiated at the bottom

```{r plotly_line_improved, exercise = TRUE}
# Line plot of per capita GDP by country
plot_ly(data = gapminder,
        x = ~ year,
        y = ~ gdpPercap,
        color = ~ country,
        colors = "#1f77b4") %>% # 1. This is the hexadecimal code for plot_ly()'s default muted blue
  add_lines(size = I(0.3)) %>% # 3. We have to use I(0.3) to make R/plot_ly() evaluate it as a single number 0.3 rather than mapping to a variable.
                               # If this doesn't make much sense to you right now it's OK. Just trust me.
  layout(showlegend = FALSE) # 2. Remove the legend

```
The `layout()` function allows you to do many customizations of the non-data portions of the graph you would have done with `theme()` in `ggplot()`.

How much did removing the legend impact your ability to read and gain insights from the chart? Maybe not as much as you think! Try rolling your mouse over each line. Because you still asked it to `color =` by country, `plot_ly()` has still implicitly grouped everything by country and displays that information when you hover over it.

## Interactive Bar Charts and Histograms

### Histograms

In `ggplot()`, you use `geom_histogram()` to specify a histogram. In `plot_ly()`, you similarly use `add_histogram()`.

```{r plotly_hist, exercise = TRUE}
# Histogram of weight of diamonds
plot_ly(data = diamonds,
        x = ~ carat) %>% 
  add_histogram()

```
### Specifying Traces: `add_XX()` vs. `add_trace()`

There are two basic ways to specify what kind of **trace** you want to add:

1. Using `add_XX()`, where XX is the trace type you want to add

2. Using the more general and flexible `add_trace()` with `type =` and possibly `mode =` arguments to indicate the *type* of trace you want to add

The two code blocks below produce identical output:

```{r plotly_line_addline}
# Line plot of per capita GDP by country
plot_ly(data = gapminder,
        x = ~ year,
        y = ~ gdpPercap,
        color = ~ country,
        colors = "#1f77b4") %>% 
  add_lines(size = I(0.3)) %>% 
  layout(showlegend = FALSE) 

```

```{r plotly_line_addtrace}
# Line plot of per capita GDP by country
plot_ly(data = gapminder,
        x = ~ year,
        y = ~ gdpPercap,
        color = ~ country,
        colors = "#1f77b4") %>% 
  add_trace(type = "scatter", # In plot_ly() speak, this line plot is actually of type "scatter" but with "lines" instead of "markers" (points) 
            mode = "lines", 
            size = I(0.3)) %>% 
  layout(showlegend = FALSE) 
```
Usually I would recommend going with method 1 (specify the desired chart type using `add_XX()`), but if you look for help online you might see things written using method 2, so I wanted you to be aware of how it might look. Some less common or more heavily customized chart types may also require you to use the more general and flexible `add_trace()`.

#### **Specifying Multiple `mode =`s**

To get certain types of plots you'll need to specify "multiple" modes by using two names with a `+` between them. This is particularly common with **traces** of `type = "scatter"`. It has three possible modes: `lines`, `markers`, and `text`. But that means there are actually *7* different types of charts you can produce with any combination of 1, 2, or all 3 of these. For example, to get lines with points we could do the following:

```{r plotly_line_addtrace2, exercise = TRUE}
# Line plot of per capita GDP by country
plot_ly(data = filter(gapminder, country %in% c("China", "India", "United States")),
        x = ~ year,
        y = ~ gdpPercap,
        color = ~ country,
        colors = "#1f77b4") %>% 
  add_trace(type = "scatter", # In plot_ly() speak, this line plot is actually of type "scatter" but with "lines" instead of "markers" (points) 
            mode = "lines+markers") %>% 
  layout(showlegend = FALSE) 
```
Note the `text` option adds text to each point instead of a marker. It doesn't really work well for this graph, but...

<mark>Challenge:</mark> For a bit of practice, try replacing the markers above with text that is the name of the country. Note to do this you will need to also specify `text = ~ <VARIABLE NAME>` to indicate the text you want at each point.

### Bar Charts

Here is where there are some differences between `ggplot()` and `plot_ly()`. In `ggplot()`, if you use `geom_bar()` R will by default also try and count the number of observations in each level of a categorical `x` variable - quite similar to a histogram. You may recall if you want it to display an actual value in the data associated with some level/value/bucket of `x`, you have to specify the `stat = "identity"` transformation.

In `plot_ly()`, `add_bars()` works like `geom_bar()` with `stat = "identity"`. In other words, it wants to map the height or length of the bar not to a count, but to a value in the data frame by default. Thus it requires you to specify what value that is as a `y` aesthetic. 

See if you can figure out how to produce an interactive bar chart of the number of diamonds in each category of cut with `plot_ly()` below. If you can't figure it out after a few minutes, you can check the solution.

```{r plotly_bar, exercise = TRUE}
# Generate counts of diamonds by cut, then plot a bar graph
diamonds2 <- diamonds %>% 
  count(cut)
diamonds2

# Bar chart of diamonds by cut
plot_ly(________) %>% 
  add_XXX()

```
```{r plotly_bar-solution}
# Generate counts of diamonds by cut, then plot a bar graph
diamonds2 <- diamonds %>% 
  count(cut)
diamonds2

# Bar chart of diamonds by cut
plot_ly(data = diamonds2,
        x = ~ cut,
        y = ~ n) %>% 
  add_bars()

```


## Advanced `plot_ly()` Techniques

### Highlighting and Subsetting Data by Users

So far we've only talked about interactivity in the form of hovering over data of interest and getting more information. But `plot_ly()` allows your users to interact with your plots in some other ways, as well. Specifically, they can select and zero in on subsets of the data. You can do this using the legend by default (if you use one), OR using the data on the plot itself with just a little added work.

#### **Highlighting With Legends**

One way to do this is using the legend. Let's go back to an earlier scatterplot of `gdpPercap` and `lifeExp`, split by continent.

```{r plotly_react1, exercise = TRUE}
# Gapminder scatterplot
plot_ly(data = gapminder,
        x = ~ gdpPercap,
        y = ~ lifeExp,
        color = ~ continent,
        colors = "Set2") %>%
  add_markers()

```
Here are some ways your, by default:

1. As we've already seen, hover over data points for more information

2. Double-click on a continent name in the legend to highlight *just* those data points. Double-click a second time to go back to showing everything.

    After the original double-click, you can single-click on other continents to gradually add them back to the plot individually, too.

3. Single-click on a continent name to *remove* that group from the plot. Single-click again to put it back.

#### **Highlighting With the Data Itself**

But what if you don't have a legend? For example, if you want someone to be able to select any 1 of your 142 countries, but you don't want to list all 142 countries in a legend? This requires just a little additional work.

Let's work with a plot of `lifeExp` over time.

```{r plotly_react2, exercise = TRUE}
# Line plot of per capita GDP by country, with highlighting capabilities
gapminder %>% 
  highlight_key(~ country) %>% # Must be the line *right before* plot_ly()
plot_ly(x = ~ year,
        y = ~ lifeExp,
        color = ~ country,
        colors = "#1f77b4") %>% 
  add_lines() %>% 
  layout(showlegend = FALSE) %>% 
  highlight()

```
Notice a few things about the above code:

* We had to pass `gapminder` through a function called `highlight_key(~ <VARIABLE NAME>)`. This functions sort of like `group_by()`; it sets the subgroups of data your user can select. In this case, we want to allow them to select any individual country to see the progression of its life expectancy.
    
    **CRITICAL**: This must be the *last* thing you write just before passing data to `plot_ly()`. Don't worry about why for now, but if you don't do this things won't work correctly!

* You also need to add `highlight()` to your `plot_ly()` code

* Your users can now, by default:

    + Single-click on a country's data point (not anywhere on the line, just near a point/year with data) to highlight that entire country's line and fade the others out.
    
    + Single-click on another country's line to switch what's highlighted
    
    + Shift+Single-click on another country's line to *add* that country to your highlighting
    
    + Double-click to remove all highlighting and return to normal
    
These behavioral triggers can be customized. Check out the `on =` and `off =` arguments in the help page for `highlight()`. Play around with them a little bit in the code above.

A more detailed accounting of everything you can do with highlighting, selecting, and user interactivity is available in [Chapter 16 of this free online book](https://plotly-r.com/client-side-linking.html#linking-animated-views), if you're interested.

<mark>Challenge:</mark> Alter the above graph so you can highlight single lines by simply placing your cursor over them, rather than clicking.

<mark>Challenge:</mark> Who are the two countries with the seemingly large life expectancy drops after 1970. Highlight them. Does your knowledge of relatively recent world history help you explain these drops? Why do we see them? What do they represent? Do a little Googling if you need.

### Modifying Tooltip Hovertext

So far we've mostly allowed `plot_ly()` to just use its default settings for what information it displays when you hover over a point with your mouse/cursor (**tooltip hovertext**). The only time we made customizations was in our original motivating example with the `midwest` data. Let's dig into this more now.

By default, `plot_ly()` shows up to 4 things as part of the hovertext, if each is applicable:

1. The value of `x`
2. The value of `y`
3. The content of a `text` parameter, if one is specified
4. The name of the **trace** if more than one is used to plot the data.

    For example, if the data is grouped by color each color is actually considered a separate trace - which is why you get the country name displayed by default next to the line plots of `gdpPercap` over time.
    
You can change this in two main ways:

* Specify the elements you want included with the `hoverinfo =` argument. By default this is `x+y+text+name`. By dropping any of these you can delete that information from the hovertext.

* Specify the exact text you want to show by altering the `text =` argument

To see these in action, let's revisit a slightly-modified version of our `midwest` example, which should make more sense now:

```{r plotly_hovertext, exercise = TRUE}
# Scatterplot of Midwestern county-level education and poverty data
plot_ly(midwest, 
        x = ~percollege, 
        y = ~percbelowpoverty) %>%
  add_markers(color = I("black"),
               text = ~paste0('</br>County: ', county, # Display "County: <COUNTY NAME>" on one line, then 
                              '</br>State: ', state),  # "State: <STATE NAME>" on the next
              hoverinfo = "text") # ONLY display the info we specify in text =
```
We used the HTML tag `</br>` to indicate we want a line break between the county name and state name. In our earlier motivating example we used a COUNTY, STATE format. Both are fine - I just wanted to show you each. It's a matter of personal preference!

<mark>Challenge:</mark> Try modifying the code below to have the hovertext contain a sensible display of the country name, continent name, and year of each data point that does not repeat any information.

```{r plotly_hovertext2, exercise = TRUE}
# Classic plot of gdpPercap vs. life expectancy
plot_ly(data = gapminder,
        x = ~ gdpPercap,
        y = ~ lifeExp,
        color = ~ country,
        colors = "#1f77b4") %>% 
  add_markers() %>% 
  layout(showlegend = FALSE) 

```
```{r plotly_hovertext2-solution}
# Classic plot of gdpPercap vs. life expectancy
plot_ly(data = gapminder,
        x = ~ gdpPercap,
        y = ~ lifeExp,
        color = ~ country,
        colors = "#1f77b4",
        text = ~paste0('Country: ', country,
                      '</br>Continent: ', continent,
                      '<br>Year: ', year),
        hoverinfo = "text") %>% 
  add_markers() %>% 
  layout(showlegend = FALSE) 

# OR

plot_ly(data = gapminder,
        x = ~ gdpPercap,
        y = ~ lifeExp,
        color = ~ country,
        colors = "#1f77b4",
        text = ~paste0('Continent: ', continent,
                      '<br> Year: ', year),
        hoverinfo = "text+name") %>% 
  add_markers() %>% 
  layout(showlegend = FALSE) 

```
A more detailed accounting of everything you can do is available in [Chapter 25 of this free online book](https://plotly-r.com/controlling-tooltips.html#tooltip-text).

### Animation

Another interactive element `plot_ly()` supports is animations (at least for traces of `type = scatter`).

The simplest way to do this is by specifying `frame = <VARIABLE NAME>` and, if necessary and appropriate, `ids = <VARIABLE NAME>` arguments. This creates a separate "frame" for each value of the indicated variable, and then cycles through those frames with a smooth transition where each value transitions from one frame to another within the indicated `ids` variable. Behold:

```{r plotly_animate, exercise = TRUE}
# Classic plot of gdpPercap vs. life expectancy
plot_ly(data = gapminder,
        x = ~ gdpPercap,
        y = ~ lifeExp,
        text = ~ paste0('Country: ', country)) %>% 
  add_markers(frame = ~ year, # Create one "frame" of the plot for each value of year
              ids = ~ country) %>%   # Causes points to transition from frame to frame within countries 
                                     # In other words, Kuwait 1952 slides to Kuwait 1957, and so on
  layout(showlegend = FALSE) 
```
You can also define frames with categorical variables. This is useful if, rather than having users play an animation, you want to allow them to select different subsets of the data to view on your plot. For example, we could create an "animation" by continent:

```{r plotly_animate2, exercise = TRUE}
# Classic plot of gdpPercap vs. life expectancy
plot_ly(data = gapminder,
        x = ~ gdpPercap,
        y = ~ lifeExp,
        text = ~ paste0('Country: ', country)) %>% 
  add_markers(frame = ~ continent, # Create one "frame" of the plot for each value of continent
              ids = ~ country) %>% # Keeping this ID prevents misleading transitions
  layout(showlegend = FALSE)
```
Your users can select the continent from which they want to see the country data.

One thing to be cautious about here is creating misleading animations. Click play on the above graph. There is still a smooth transition, but the points don't "slide" from one spot to another like they did above. This is good! Countries don't exist on more than one continent in the way they do in more than one year.

<mark>Challenge:</mark> But try deleting the `ids =` argument above. How does the plot animate now? Think about whether and how this could be misleading.

<mark> Challenge:</mark> An alternative even if you deleted the `ids =` argument above is to specify you want `plot_ly()` to do something other than its default smooth transitions. You can accomplish this by using the code `animation_opts(transition = 0)` as a new line. (`transition = 0` simply removes any transitions.) Try it above! Do you like this any better than the original? Again, it's a personal/stylistic preference!

If you're interested, for far more details on making animations in `plot_ly()` you can read [Chapter 14 of this free online textbook](https://plotly-r.com/animating-views.html). 

You can also do animations in `ggplot()` using the `gganimate` package! [Here](https://gganimate.com/articles/gganimate.html) is a very basic introduction. [Here](https://www.datanovia.com/en/blog/gganimate-how-to-create-plots-with-beautiful-animation-in-r/) are some more examples, including how to create an animation that "draws" a line graph live by gradually revealing more and more data over time! And [here](http://rstudio-pubs-static.s3.amazonaws.com/494188_154898728be3411ebab39050ca8a8dcd.html) is an example of how to animate a football play in R using `gganimate`!

## Summary

Congratulations! You now know a little more than the basics - though still aren't an expert in - `plot_ly()`. Now you can create graphics your users can interact with, rather than mere static plots. Pretty cool, right?

We've barely scratched the surface of what we can do with `plotly`. For more information I would recommend Googling specific thigns you want to do or taking a look at [this free online book](https://plotly-r.com/index.html). Some of it is likely too advanced for some of you, but it should help with the basics. Plus it's the definitive bible of `plotly` so I'm including it here.

Up next, you're going to learn how to build interactive apps in R using `shiny`. So more neat stuff is on the horizon before we get into the vegetables of data cleaning.